---
title: "Team Essay 1"
author: "Dhruvi, Vi, Huy, Alex, Jesse, Harshal"
date: "2/16/2021"
output: pdf_document
---

```{r}
library("readxl")
library("tidyverse")
```

### Dataset Information

#### National Employement Male vs Female - unemployment.xlsx
+ X = national unemployment rate for adult males
+ Y = national unemployment rate for adult females

#### Deaths caused by Firearms - firearms.xlsx
The data gives the number of deaths caused by firearms in Australia from 1983 to 1997, expressed as rate per 100,000 of population.
+ Year = Year
+ Rate = Number of deaths caused by firearms per 100,000 population

**Importing the data file**

```{r}
unemp_dat <- read_excel("unemployment.xlsx")
unemp_dat <- unemp_dat %>% 
  rename(
    "mRate" = "X",
    "fRate" = "Y")
unemp_data <- unemp_dat[1:9, ]
unemp_data

firearms <- read_excel("firearms.xlsx")
firearms_data <- firearms[1:11, ]
firearms_data
```
 
**Formula and basics**

Linear regression: `y = b0 + b1*x + e`

+ *b0* is the intercept of regression line: predicted value when *x = 0*.
+ *b1* is the slope of the regression line.
+ *e* is the error term, the part of y that can be explained through the regression model.

### Visualization
```{r}
ggplot(unemp_data, aes(x = mRate, y = fRate)) +
  geom_point() +
  stat_smooth()

ggplot(firearms_data, aes(x = Year, y = Rate)) +
  geom_point() +
  stat_smooth() 
```

```{r}
cor(unemp_data$fRate, unemp_data$mRate)

cor(firearms_data$Rate, firearms_data$Year)
```

### Computation

+ model1: `fRate = b0 + b1 * mRate`
+ model2: `rate = b0 + b1 * Year`

```{r}
model1 <- lm(fRate ~ mRate, data = unemp_data)
model1

model2 <- lm(Rate ~ Year, data = firearms_data)
model2
```

**Interpretation of Model 1: unemployment data**

From the output above:

+ the estimated regression line equation can be written as follow: `fRate = 2.187 + (0.608) * mRate`

+ the intercept (b0) is 2.187.

+ the regression beta coefficient for the variable pressure (b1), also known as the slope, is 0.608.

**Interpretation of Model 2: firearms data**

From the output above:

+ the estimated regression line equation can be written as follow: `rate = 273.416 + (-0.136) * Year`

+ the intercept (b0) is 273.416.

+ the regression beta coefficient for the variable youtube (b1), also known as the slope, is -0.136.

### Estimates

**Unemployement Data**

+ Model: `fRate = 2.187 + (0.608) * mRate`

1. mRate = 5.1
  + Estimate of fRate: 5.3
  + Actual: 4.2
  
2. mRate = 4.7
  + Estimate of fRate: 5.0
  + Actual: 4.0
  
3. mRate = 4.4
  + Estimate of fRate: 4.9
  + Actual: 4.4
  
4. mRate = 5.8
  + Estimate of fRate: 5.7
  + Actual: 5.2

**Firearms Data**

+ Model: `rate = 273.416 + (-0.136) * Year`

1. Year = 1994
  + Estimate: 2.23
  + Actual: 2.95
  
2. Year = 1995
  + Estimate: 2.096
  + Actual: 2.72
  
3. Year = 1996
  + Estimate: 1.96
  + Actual: 2.96
  
4. Year = 1997
  + Estimate: 1.82
  + Actual: 2.30

### Regression line

```{r}
ggplot(unemp_data, aes(x = mRate, y = fRate)) +
  geom_point() +
  stat_smooth(method = lm)

ggplot(firearms_data, aes(x = Year, y = Rate)) +
  geom_point() +
  stat_smooth(method = lm) 
```

### Model Summary

```{r}
summary(model1)

summary(model2)
```
