---
title: 'Essay 5: Principal Component Analysis and Clustering'
author: "Dhruvi, Vi, Huy, Harshal"
date: "4/23/2021"
output: html_document
---

## (1) INTRODUCTION

### Principal Component Analysis and K-means Clustering


### Loading R packages

```{r, results=FALSE, warning=FALSE, message=FALSE}
library("dplyr")
library("tidyverse") #for data manipulation and visualization
library("corrplot")
library("animation")
library("cluster")
library("GGally")
library("factoextra")
library("ggcorrplot")
```

## (2) DATA DESCRIPTION

### Examples of data and problem

**Wine Data set**

These data are the results of a chemical analysis of wines grown in the same region in Italy but derived from three different cultivars. The analysis determined the quantities of 13 constituents found in each of the three types of wines. The dataset contains 13 attributes with 178 rows.

1) Alcohol
2) Malic acid
3) Ash
4) Alcalinity of ash
5) Magnesium
6) Total phenols
7) Flavanoids
8) Nonflavanoid phenols
9) Proanthocyanins
10) Color intensity
11) Hue
12) OD280/OD315 of diluted wines
13) Proline

*Our goal is to try to group similar observations together and determine the number of possible clusters.*

**Importing the data file**

```{r, echo=TRUE, results='hide', warning=FALSE, message=FALSE}

wine.cols = c("Alcohol", "Malic_Acid", "Ash", "Ash_Alcanity", "magnesium", "Total_Phenols", "Flavanoids",	"Nonflavanoid_Phenols", "Proanthocyanins", "Color_Intensity", "Hue", "OD280", "Proline")

wine <- read.table("wine.data", sep = ",", row.names=NULL, col.names=wine.cols, nrows=178)[ ,2:14]

wine <- wine %>% 
  drop_na() %>% #remove NA values
  distinct() # remove duplicated rows

summary(wine)
```

### Visualization

First we plot the histogram of each attribute. The idea here is to visualize the data distribution for each feature.

```{r}
# Histogram for each Attribute
wine %>%
  gather(Attributes, value, 1:13) %>%
  ggplot(aes(x=value, fill=Attributes)) +
  geom_histogram(colour="black", show.legend=FALSE) +
  facet_wrap(~Attributes, scales="free_x") +
  labs(x="Values", y="Frequency",
       title="Wines Attributes - Histograms") +
  theme_bw()
```

```{r}
wine %>%
  gather(Attributes, values) %>%
  ggplot(aes(x=reorder(Attributes, values, FUN=median), y=values, fill=Attributes)) +
  geom_boxplot(show.legend=FALSE) +
  facet_wrap(~Attributes, scales="free", nrow = 4) +
  labs(title="Wines Attributes - Boxplots") +
  theme_bw() +
  theme(axis.title.y=element_blank(),
        axis.title.x=element_blank()) 
```

```{r}
ggcorr(wine,
    nbreaks = 10,
    label = TRUE,
    label_size = 3,
    color = "grey50")
```


**Data Normalization**

We need to normalize the data to express it in the same range of values.

```{r}
winesNorm <- as.data.frame(scale(wine))

summary(winesNorm)
```

## (3) ANALYSIS

### Computation

```{r}
distance <- get_dist(winesNorm) 
fviz_dist(distance, gradient = list(low = "#00AFBB", mid = "white", high = "#FC4E07")) + 
  theme(
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank()
  )
```

```{r}
k2 <- kmeans(winesNorm, centers = 3, nstart = 25)

k2

fviz_cluster(k2, data = winesNorm)
```

```{r}
set.seed(123)

# function to compute total within-cluster sum of square 
wss <- function(k) {
  kmeans(winesNorm, k, nstart = 10 )$tot.withinss
}

# Compute and plot wss for k = 1 to k = 15
k.values <- 1:15

# extract wss for 2-15 clusters
wss_values <- map_dbl(k.values, wss)

plot(k.values, wss_values,
       type="b", pch = 19, frame = FALSE, 
       xlab="Number of clusters K",
       ylab="Total within-clusters sum of squares")
```

```{r}
set.seed(123)

fviz_nbclust(winesNorm, kmeans, method = "wss")
```


### Interpretation of Model: 


  
## (4) MODEL EVALUATION

### Regression line


### Model Assessment 



### Model Summary



### Coefficients significance



### Prediction and Model accuracy


## (5) CONCLUSION

### Summary



## (6) REFERENCES

https://stat.ethz.ch/R-manual/R-patched/library/datasets/html/USArrests.html

https://www.kaggle.com/xvivancos/tutorial-clustering-wines-with-k-means

https://www.kaggle.com/rodrigofragoso/explained-k-means-pca-visualization

https://uc-r.github.io/kmeans_clustering

https://www.guru99.com/r-k-means-clustering.html